The importance of transient resonances in extreme-mass-ratio inspirals (new title?)

### 2015-05-30 res_char.tex
Location of resonances

\author{Robert H. Cole}
\email[]{rhc26@ast.cam.ac.uk}
\affiliation{Institute of Astronomy, Madingley Road, Cambridge, CB3 0HA, United Kingdom}
\author{Christopher P.L. Berry}
\email[]{cplb@star.sr.bham.ac.uk}
\affiliation{Institute of Astronomy, Madingley Road, Cambridge, CB3 0HA, United Kingdom}
\affiliation{School of Physics and Astronomy, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom}
\author{Priscilla Ca\~{n}izares}
\email[]{pcm@ast.cam.ac.uk}
\author{Jonathan R. Gair}
\email[]{jgair@ast.cam.ac.uk}
\affiliation{Institute of Astronomy, Madingley Road, Cambridge, CB3 0HA, United Kingdom}

@@

\begin{abstract}
\emph{Update abstract}
The inspiral of stellar-mass compact objects into supermassive black holes provides a wealth of information about the strong gravitational-field regime via the emission of gravitational waves. In order to detect such signals and extract the desired information, it is necessary to possess a bank of accurate waveform templates. For computational efficiency, adiabatic templates are often used, which accurately reproduce orbit-averaged trajectories arising from the first order dissipative gravitational self-force. Other effects are however neglected, in particular that of transient resonances, where the radial and poloidal fundamental frequencies become commensurate. During such resonances, the flux of gravitational waves can be diminished or enhanced, leading to a shift in the compact-object trajectory and, ultimately, the phase of the emitted radiation. In this work, an astrophysical population of detectable extreme-mass-ratio inspirals is studied and it is found that a large proportion shall encounter a low-order resonance in the later stages of inspiral. The resulting effect on signal-to-noise recovery is small due to the low eccentricity of the studied population, but nevertheless at least one fifth of the systems may become undetectable if transient resonances aren't accounted for.
\end{abstract}

@@

\emph{Move location stuff to appendix ... the bits here need tweaking to reflect that.}

The first step in studying the effect of transient resonances is to locate orbital parameters for which the frequencies are commensurate. We can calculate the frequencies and so we are left with the problem of solving $\Omega = n_r \Omega_r - n_\theta \Omega_\theta = 0$ numerically. \Figref{res-plane-2-5-95} shows the semilatus rectum, eccentricity and (cosine of the) inclination angle of the $\nu = 2/5$ resonance surface for a BH of spin $a_\ast = 0.95$. 
\begin{figure}[htbp]
\centering
\includegraphics[width=0.46\textwidth]{Fig_res-2-5-95-plane}
\caption{\label{fig:res-plane-2-5-95}Location of the $2/5$ resonance surface for an $a_\ast = 0.95$ BH in terms of orbital semilatus rectum $p$, eccentricity $e$ and inclination $\iota$.}
\end{figure}
The resonances occur at relatively small periapses, corresponding to regions of strong-field gravity.
Similar resonance surfaces can be found for other spin values and for other resonances. When considering the full parameter set of $\{p,e,\iota,a_\ast,\nu\}$, it is apparent that the search for resonances becomes expensive as a consequence of the dimensionality. It is therefore useful to have a guide of where to look. We shall attempt to build a model for the resonance plane.

@@

\subsection{Resonant flux enhancement}

@@

\section{Location of resonances}
Brink, Geyer and Hinderer~\cite{Brink2013} provide series expansions for the location of resonances in the limit of equatorial orbits for small spin and eccentricity. We do not follow this approach of trying to find analytic expressions for the resonance surface. The expressions become complicated when venturing away from limiting cases. Instead, we build an approximate phenomenological model and fit this to the resonance plane. %\footnote{A stopped clock is precisely correct twice a day, while a clock that is five minutes slow is never right but will often be close enough. Within this metaphor, our approximation would be a clock that varies between telling the correct time and being fifteen minutes off throughout the day: it is generally going to be good enough to give you a sense of the time but if you have an important event you should check with something better.}
This should be useful for designating the region in which resonance could be expected. To locate them precisely, it is necessary to solve $\Omega = 0$ numerically; the approximate model should give a suitable starting point.

>>>>

\author{Robert H. Cole}
\email[]{rhc26@ast.cam.ac.uk}
\affiliation{Institute of Astronomy, Madingley Road, Cambridge, CB3 0HA, United Kingdom}
\author{Christopher P.L. Berry}
\email[]{cplb@star.sr.bham.ac.uk}
\affiliation{School of Physics and Astronomy, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom}
\affiliation{Institute of Astronomy, Madingley Road, Cambridge, CB3 0HA, United Kingdom}
\author{Priscilla Ca\~{n}izares}
\email[]{pcm@ast.cam.ac.uk}
\author{Jonathan R. Gair}
\email[]{jgair@ast.cam.ac.uk}
\affiliation{Institute of Astronomy, Madingley Road, Cambridge, CB3 0HA, United Kingdom}

@@

\begin{abstract}
\emph{Update abstract}
The inspiral of stellar-mass compact objects into supermassive black holes provides a wealth of information about the strong gravitational-field regime via the emission of gravitational waves. In order to detect such signals and extract the desired information, it is necessary to possess a bank of accurate waveform templates. For computational efficiency, adiabatic templates are often used, which accurately reproduce orbit-averaged trajectories arising from the first order dissipative gravitational self-force. Other effects are however neglected, in particular that of transient resonances, where the radial and poloidal fundamental frequencies become commensurate. During such resonances, the flux of gravitational waves can be diminished or enhanced, leading to a shift in the compact-object trajectory and, ultimately, the phase of the emitted radiation. In this work, an astrophysical population of detectable extreme-mass-ratio inspirals is studied and it is found that a large proportion shall encounter a low-order resonance in the later stages of inspiral. The resulting effect on signal-to-noise recovery is small due to the low eccentricity of the studied population, but nevertheless at least one fifth of the systems may become undetectable if transient resonances are not accounted for.
\end{abstract}

@@

\emph{Move location stuff to appendix. Bits here have been tweaked.}

The first step in studying the effect of transient resonances is to locate orbital parameters for which the frequencies are commensurate. We can calculate the frequencies and so we are left with the problem of solving $\Omega = n_r \Omega_r - n_\theta \Omega_\theta = 0$ numerically. When considering the full parameter set of $\{p,e,\iota,a_\ast,\nu\}$, it is apparent that the search for resonances becomes expensive as a consequence of the dimensionality. It is therefore useful to have a guide of where to look. In \apref{location} we build a simple approximate model as a starting point for the numerical search. The resonances occur at relatively small periapses, corresponding to regions of strong-field gravity. Having located where in an inspiral we can expect to encounter a transient resonance, we must now consider its impact. In \secref{res-time} we determine the characteristic time-scales describing resonance, and in \secref{flux-enhance} we look at the impact of passing through a resonance on the evolution of the orbit.

@@

\subsection{Resonant flux enhancement}\label{sec:flux-enhance}

@@

\section{Location of resonances}\label{sec:location}

We can find the location of resonances by numerically solving $\Omega = n_r \Omega_r - n_\theta \Omega_\theta = 0$. \Figref{res-plane-2-5-95} shows the semilatus rectum, eccentricity and (cosine of the) inclination angle of the $\nu = 2/5$ resonance surface for a BH of spin $a_\ast = 0.95$. 
\begin{figure}[htbp]
\centering
\includegraphics[width=0.46\textwidth]{Fig_res-2-5-95-plane}
\caption{\label{fig:res-plane-2-5-95}Location of the $2/5$ resonance surface for an $a_\ast = 0.95$ BH in terms of orbital semilatus rectum $p$, eccentricity $e$ and inclination $\iota$.}
\end{figure}
This is almost planar, inspiring us to look for a simple description that can help guide our search for resonance locations. Brink, Geyer and Hinderer~\cite{Brink2013} provide series expansions for the location of resonances in the limit of equatorial orbits for small spin and eccentricity. We do not follow this approach of trying to find analytic expressions for the resonance surface. The expressions become complicated when venturing away from limiting cases. Instead, we build an approximate phenomenological model and fit this to the resonance plane. %\footnote{A stopped clock is precisely correct twice a day, while a clock that is five minutes slow is never right but will often be close enough. Within this metaphor, our approximation would be a clock that varies between telling the correct time and being fifteen minutes off throughout the day: it is generally going to be good enough to give you a sense of the time but if you have an important event you should check with something better.}
This should be useful for designating the region in which resonance could be expected. To locate them precisely, it is necessary to solve $\Omega = 0$ numerically; the approximate model gives a suitable starting point.
